Higher-Order unification for free!

### 2024-04-24 paper/main.tex
relecture of tex

\noindent
The proof found by the solver back-chains on rule 3 (the only rule
about the $\forall$ quantifier), and then solves the premises with
rules 1 and 2 respectively.
Note that rule 3 features a second-order parameter \coqIn{P} that represents
a function of type \coqIn{A ~$\to$~ Prop} (a 1-argument predicate over \coqIn{A}).
The solver has to infer a value for \coqIn{P} by unifying the conclusion
of rule 3 with the goal, and in particular, it has to solve the unification
problem \coqIn{P x = nfact x 3}. This higher-order problem falls in the so-called 
pattern fragment and admits a unique
solution $\customlabel{solution:intro}{\rho}$ that assigns
the term \elpiIn{~$\lambda$~x.nfact x 3} to \coqIn{P}.
In the rest of this paper shall refer to the pattern fragment as
\llambda~\cite{miller92jsc}.

In order to implement such a search in Elpi, we shall describe the encoding
of Coq terms and then the encoding of instances as rules.
Elpi comes equipped with
an Higher Order Abstract Syntax (HOAS~\cite{10.1145/53990.54010}) datatype of
terms, called \elpiIn{tm}, that includes (among others) the following
constructors:

@@

\noindent
The root cause of the problems we outlined in this example
is a subtle mismatch between the equational theories of the meta language
and the object language, which in turn makes the 
unification procedures of the meta language weak.
The equational theory of the meta language Elpi encompasses
$\eta\beta$-equivalence and its unification procedure can solve higher-order
problems in the pattern fragment. Although 
the equational theory of Coq is much richer, for efficiency and predictability
reasons, automatic proof search procedures typically employ a unification
procedure that only captures a $\eta\beta$-equivalence and only operates
in \llambda. The similarity is striking, but one needs to exercise some caution in order to
simulate a higher-order logic program in Coq using the unification of Elpi.

\paragraph{Contributions}
In this paper we identify a minimal language \Fo{} in which the problems
sketched in the introduction can be formally described.
We detail an encoding of a logic program in \Fo{} to a strongly related
logic program in \Ho (the language of the meta-language) and we show that
the higher-order unification procedure of the meta language \Ue{} can be
efficiently used to simulate a higher-order unification procedure \Uo for
the object language that features $\eta\beta$-conversion. We show how \Uo
can be extended with heuristics to deal with problems outside the pattern
fragment.\\
Section~\ref{sec:problem-statement} states the problem formally and gives the
intuition behind our solution; \cref{sec:simulation} sets up a basic
simulation of first-order logic programs, \cref{sec:llam} and \cref{sec:eta}
extend it to higher-order logic programs in the pattern fragment
while \cref{sec:beta} goes beyond the pattern fragment.
Section \ref{sec:implementation} discusses the implementation in Elpi.\\
The $\lambda$Prolog code discussed in the paper can be accessed at the
address \githubUrl.

@@

\noindent
Note that the symbol \elpiIn{|} separates the head of a list from the tail,
that \elpiIn{forall2} applies a binary predicate to two lists (of the same
length). Finally \elpiIn{=>} is $\lambda$Prolog syntax to augment the
current program with an extra rule, and both procedures use it to
assume that the fresh nominal constants introduces by \elpiIn{pi} are
equal to themselves.

@@

\begin{proposition}[Good unification]\label{prop:good-unif}
A good unification $\UnifRel\;$ for equality $\EqualRel\;$ in the domain
$\mathcal{D}$ satisfies the following properties:
  \begin{gather}
    \unifcorrectX{\EqualRel}{\UnifRel}{t}{\{~t_1,~ t_2~\} \subseteq \mathcal{D}}\\
    \unifcompleteX{\EqualRel}{\UnifRel}{t}{\{~ t_1,~ t_2 ~\} \subseteq \mathcal{D}}
  \end{gather}
\end{proposition}

@@

\subsection{The solution (in a nutshell)}
\label{sec:nutshell}
A term $s$ is compiled to a term $t$ where every
``problematic'' sub term $p$ is replaced by a fresh unification variable $h$
with an accessory \emph{link} that represents a suspended unification problem
$h \Ue p$. As a result \Ue is ``well behaved'' on $t$, in the sense that
it does not contradict \Eo as it would otherwise do on the
``problematic'' sub-terms.\\
We now define ``problematic'' and ``well behaved'' more formally
We use the $\Diamond$ symbol since it stands for ``possibly'' in modal logic
and all problematic terms are characterized by some ``uncertainty''.

@@

\noindent
An example of a \maybebeta{} term is the application $F \appsep x$.
This term is problematic since the application node of
its syntax tree cannot be used to justify a
unification failure, i.e. by properly instantiating $F$ the term
head constructor may become a $\lambda$, or a constant, or remain an application.

@@

\noindent
An example of a term $s$ in \maybeeta{} is
$\lambda x.\lambda y.F \appsep y \appsep x$
since the substitution
$\rho = \{ F \mapsto \lambda a.\lambda b.f \appsep b \appsep a\}$
makes $\rho s = \lambda x.\lambda y.f \appsep x \appsep y$,
which is the eta long form of $f$. This term is problematic since
its leading $\lambda$ abstraction cannot justify a
unification failure against a constant $f$.

@@

\noindent
We write $\wellb(t)$ as a short for $\wellb(\{t\})$.
We claim our compiler validates the following property:

@@

\noindent
A solution to a \Fo{} variable is a plain term, that is \elpiIn{fsubst}
is an abbreviation for \elpiIn{mem fm}.
%We call \elpiIn{fsubst} the memory of \Fo{}, while we call \elpiIn{subst}
%the one of \Ho.

@@

The signature of the \elpiIn{comp} predicate below allows for the generation of
links (suspended unification problems), which play no role in this section
but play a major role in \cref{sec:llam,sec:eta,sec:beta}.
With respect to \cref{sec:problem-statement}, the signature also allows
for updates to the substitution.

@@

\subsubsection{Occur check}\label{sec:oc}
Since compilation moves problematic terms out of the sight of \Ue{},
that procedure can only perform a partial occur check. For example, the
unification problem $X \Ue f~Y$ cannot generate a cyclic substitution alone,
but should be disallowed if a $\linkStore$ contains a link like
$\linketaM{}{Y}{\lambda z.X_z}$: we don't know yet if $Y$ will feature
a lambda in head position, but we surely know it contains $X$, hence
$f~Y$ and that fails the occur check.
The procedure \elpiIn{occur-check-links} is in charge of
performing this check that is needed in order to
guarantee \fullref{prop:fidelity}.

@@

The final step is to decompile each assignment. The \elpiIn{val}
node carries a term that easy to decompile since \mapStore is a bijection.
Since the \elpiIn{flam} node
carries no additional information (other than the function body),
each \elpiIn{abs} node can be decompiled to a \elpiIn{flam} one.

@@

We already have all the pieces to show the code of \Ue{}.

@@

\begin{lemma}[Properties of \Uo in \wellb]\label{prop:uniffow}
  The implementation of \Uo above is a good unification
  for \Eo
  (as per \cref{prop:good-unif})
  in the domain \wellb.
\end{lemma}
\begin{proof}[Proof sketch]
 In this setting, \Ee is as strong as \Eo on ground terms. What we have to show is that whenever two different \Fo{}
terms can be made equal by a substitution $\rho$ (plus the \ref{clause:beta1}
and \ref{clause:beta2} if needed), we can produce this $\rho$ by finding
a $\sigma$ via \Ue{} on the corresponding \Ho terms and by decompiling it.
If we look at the syntax of \Fo{}, terms the only interesting case is
\elpiIn{fuva X ~\Uo~s}. In this case, after compilation, we have
$Y \Ue t$ that succeeds with $\sigma = \{ Y \mapsto t\}$ and
$\sigma$ is decompiled to $\rho = \{ X \mapsto s\}$ by \cref{prop:comprt}.
\end{proof}

@@

\begin{theorem}[Fidelity in $\wellb_\beta$] \Fullref{prop:simulation} and
\fullref{prop:fidelity} hold if $\;\wellb_\beta(\foUnifPb)$.
\label{th:fidelity-beta}
\end{theorem}
\begin{proof}[Proof sketch] thanks to \cref{lem:w-enforcement-maybebeta}
  \Ue is as powerful as \Uo in \maybebeta, as well as in \wellb by
  \cref{prop:uniffowb}.
\end{proof}

@@

\begin{proof}[Proof sketch]\todo{to reformulate}
  % For any step $i$, we want to prove that $\fstep_i$ succeeds iff $\hstep_i$
  % succeeds. A $\hstep_i$ is made by a unification
  % $u = \sigma_{i-1}\hoUnifPb_{i_l} \Ue \sigma_{i-1}\hoUnifPb_{i_r} \mapsto \sigma_i'$ and a link
  % progression $p = \progress(\linkStore_{i-1},\sigma_i') \mapsto (\linkStore_i,\sigma_i)$.
  We want to prove that \fstep succeeds iff \hstep
  succeeds, where \hstep is made by a unification step $u$ and a link
  progression $p$.
  Without loss of generality we consider a
  \Fo unification problem of the form $s_1 \Uo s_2$
  where $s_1 \in \maybeeta$ and is compiled to a variable $X$ with the accessory link
  $\linketaM{\Gamma}{X}{\lambda x.t_1}$. 
  The unification $u$ always succeeds, since $X$ is a fresh variable, we 
  only have to consider the link progression $p$.
  Two cases should be analyzed:
  \begin{itemize}
    \item $s_2 \notin \maybeeta$,
          in this case $s_2$ is compilet to $t_2$ and the unification step $u$ assigns $t_2$ to $X$
          and triggers \progressetaleft if $t_2$ is rigid.
          If $s_1$ and $s_2$ unify thanks to \ref{clause:eta1} then
          progression $p$ unifies $\lambda x.t_1$ with the $\eta$-expansion of $t_2$ namly
          $\lambda x.t_2\appsep x$, otherwise they unify thanks to \cref{clause:lam-lam}\todo{name}
          and the link progress $p$ unifies $\lambda x.t_1$ with $t_2$.\\ Alternatively, if $s_1$ and $s_2$ fail
          to unify, it cannot be because of the head $\lambda$ constructors,
          since with \cref{lam-lam}\todo{name} or \ref{clause:eta1} move under it and find
          different sub terms $s_1'$ and $s_2'$. By \cref{th:fidelity-beta},
          \Ue fails on terms $t_1'$ and $t_2'$ even if \progressetaleft
          $\eta$-expands $t_2$.

@@

\begin{elpicode}
link-eta L R :- var L, maybe-eta R HolesInR, !,
  declare_constraint (link-eta L R) [L|HolesInR].
link-eta L R :- not (var L), !, eta-progress-lhs L R.
link-eta L R :- not (maybe-eta R), !, contract R S, L = S.
\end{elpicode}
\todo{check progress-rhs is back}

>>>>

\noindent
The proof found by the solver back-chains on rule 3 (the only rule
about the $\forall$ quantifier), and then solves the premises with
rules 1 and 2 respectively.
Note that rule 3 features a second-order parameter \coqIn{P} that represents
a function of type \coqIn{A ~$\to$~ Prop} (a 1-argument predicate over \coqIn{A}).
The solver has to infer a value for \coqIn{P} by unifying the conclusion
of rule 3 with the goal, and in particular, it has to solve the unification
problem \coqIn{P x = nfact x 3}. This higher-order problem falls in the so-called 
pattern fragment and admits a unique
solution $\customlabel{solution:intro}{\rho}$ that assigns
the term \elpiIn{~$\lambda$~x.nfact x 3} to \coqIn{P}.
In the rest of this paper, we refer to the pattern fragment as
\llambda~\cite{miller92jsc}.

In order to implement such a search in Elpi, we shall describe the encoding
of Coq terms and then the encoding of instances as rules.
Elpi comes equipped with
an Higher-Order Abstract Syntax (HOAS~\cite{10.1145/53990.54010}) datatype of
terms, called \elpiIn{tm}, that includes (among others) the following
constructors:

@@

\noindent
The root cause of the problems we outlined in this example
is a subtle mismatch between the equational theories of the meta language
and the object language, which in turn makes the 
unification procedures of the meta language weak.
The equational theory of the meta-language Elpi encompasses
$\eta\beta$-equivalence and its unification procedure can solve higher-order
problems in the pattern fragment. Although 
the equational theory of Coq is much richer, for efficiency and predictability
reasons, automatic proof search procedures typically employ a unification
procedure that only captures a $\eta\beta$-equivalence and only operates
in \llambda. The similarity is striking, but one needs to exercise some caution in order to
simulate a higher-order logic program in Coq using the unification of Elpi.

\paragraph{Contributions}
In this paper we identify a minimal language \Fo{} in which the problems
sketched in the introduction can be formally described.
We detail an encoding of a logic program in \Fo{} to a strongly related
logic program in \Ho (the language of the meta language) and we show that
the higher-order unification procedure of the meta language \Ue{} can be
efficiently used to simulate a higher-order unification procedure \Uo for
the object language that features $\eta\beta$-conversion. We show how \Uo
can be extended with heuristics to deal with problems outside the pattern
fragment.\\
Section~\ref{sec:problem-statement} states the problem formally and gives the
intuition behind our solution; \cref{sec:simulation} sets up a basic
simulation of first-order logic programs, \cref{sec:llam} and \cref{sec:eta}
extend it to higher-order logic programs in the pattern fragment
while \cref{sec:beta} goes beyond the pattern fragment.
Section \ref{sec:implementation} discusses the implementation in Elpi.\\
The $\lambda$Prolog code discussed in the paper can be accessed at the
address \githubUrl.

@@

\noindent
Note that the symbol \elpiIn{|} separates the head of a list from the tail,
that \elpiIn{forall2} applies a binary predicate to two lists (of same
length). Finally \elpiIn{=>} is $\lambda$Prolog syntax to augment the
current program with an extra rule, and both procedures use it to
assume that the fresh nominal constants introduces by \elpiIn{pi} are
equal to themselves.

@@

\begin{proposition}[Good unification]\label{prop:good-unif}
A good unification $\UnifRel\;$ for equality $\EqualRel\;$ in the domain
$\mathcal{D}$ satisfies the following properties:\noteD{use $\sigma$, not $\rho$}
  \begin{gather}
    \unifcorrectX{\EqualRel}{\UnifRel}{t}{\{~t_1,~ t_2~\} \subseteq \mathcal{D}}\\
    \unifcompleteX{\EqualRel}{\UnifRel}{t}{\{~ t_1,~ t_2 ~\} \subseteq \mathcal{D}}
  \end{gather}
\end{proposition}

@@

\subsection{The solution (in a nutshell)}
\label{sec:nutshell}
A term $s$ is compiled to a term $t$ where every
``problematic'' subterm $p$ is replaced by a fresh unification variable $h$
with an accessory \emph{link} that represents a suspended unification problem
$h \Ue p$. As a result, \Ue is ``well behaved'' on $t$, in the sense that
it does not contradict \Eo as it would otherwise do on the
``problematic'' sub-terms.\\
We now define ``problematic'' and ``well behaved'' more formally
We use the $\Diamond$ symbol since it stands for ``possibly'' in modal logic
and all problematic terms are characterized by some ``uncertainty''.

@@

\noindent
An example of a \maybebeta{} term is the application $F \appsep x$.
This term is problematic since the application node of 
its syntax tree cannot be used to justify a
unification failure, i.e. by properly instantiating $F$, the term\noteD{or a name}
head constructor may become a $\lambda$, or a constant, or remain an application.

@@

\noindent
An example of a term $s$ in \maybeeta{} is
$\lambda x.\lambda y.F \appsep y \appsep x$
since the substitution
$\rho = \{ F \mapsto \lambda a.\lambda b.f \appsep b \appsep a\}$\noteD{a,b are reserved for constants}
makes $\rho s = \lambda x.\lambda y.f \appsep x \appsep y$,
which is the eta long form of $f$. This term is problematic since
its leading $\lambda$ abstraction cannot justify a
unification failure against a constant $f$.

@@

\noindent
We write $\wellb(t)$ as a short for $\wellb(\{t\})$.
We claim our compiler validates the following properties:

@@

\noindent
A solution to a \Fo{} variable is a plain term, that is, \elpiIn{fsubst}
is an abbreviation for \elpiIn{mem fm}.
%We call \elpiIn{fsubst} the memory of \Fo{}, while we call \elpiIn{subst}
%the one of \Ho.

@@

The signature of the \elpiIn{comp} predicate below allows for the generation of
links (suspended unification problems), which plays no role in this section
but plays a major role in \cref{sec:llam,sec:eta,sec:beta}.
With respect to \cref{sec:problem-statement}, the signature also allows
for updates to the substitution.

@@

\subsubsection{Occur check}\label{sec:oc}
Since compilation moves problematic terms out of the sight of \Ue{},
that procedure can only perform a partial occur check. For example, the
unification problem $X \Ue f~Y$ cannot generate a cyclic substitution alone,
but should be disallowed if $\linkStore$ contains a link like
$\linketaM{}{Y}{\lambda z.X_z}$: we don't know yet if $Y$ will feature
a lambda in head position, but we surely know it contains $X$, hence,
% $f~Y$ and 
that fails the occur check.
The procedure \elpiIn{occur-check-links} is in charge of
performing this check that is needed in order to
guarantee \fullref{prop:fidelity}.

@@

The final step is to decompile each assignment. The \elpiIn{val}
node carries a term that is easy to decompile since \mapStore is a bijection.
Since the \elpiIn{flam} node
carries no additional information (other than the function body),
each \elpiIn{abs} node can be decompiled to a \elpiIn{flam} one.

@@

We already have all the ingredients to show the code of \Uo{}.

@@

\begin{lemma}[Properties of \Uo in \wellb]\label{prop:uniffow}
  The implementation of \Uo above is a good unification
  for \Eo
  (as per \cref{prop:good-unif})
  in the domain \wellb.
\end{lemma}
\begin{proof}[Proof sketch]
 In this setting, \Ee is as strong as \Eo on ground terms\noteD{module var reflexivity}. What we have to show is that whenever two different \Fo{}
terms can be made equal by a substitution $\rho$ (plus the \ref{clause:beta1}
and \ref{clause:beta2} if needed), we can produce this $\rho$ by finding
a $\sigma$ via \Ue{} on the corresponding \Ho terms and by decompiling it.
If we look at the syntax of \Fo{}, terms the only interesting case is
\elpiIn{fuva X ~\Uo~s}. In this case, after compilation, we have
$Y \Ue t$\noteD{shouldn't be uva Y [] \Ue t} that succeeds with $\sigma = \{ Y \mapsto t\}$ and
$\sigma$ is decompiled to $\rho = \{ X \mapsto s\}$ by \cref{prop:comprt}.
\end{proof}

@@

\begin{theorem}[Fidelity in $\wellb_\beta$] \Fullref{prop:simulation} and
\fullref{prop:fidelity} hold if $\;\wellb_\beta(\foUnifPb)$.
\label{th:fidelity-beta}
\end{theorem}
\begin{proof}[Proof sketch] Thanks to \cref{lem:w-enforcement-maybebeta}
  \Ue is as powerful as \Uo in \maybebeta, as well as in \wellb by
  \cref{prop:uniffowb}.
\end{proof}

@@

\begin{proof}[Proof sketch]\todo{to reformulate}
  % For any step $i$, we want to prove that $\fstep_i$ succeeds iff $\hstep_i$
  % succeeds. A $\hstep_i$ is made by a unification
  % $u = \sigma_{i-1}\hoUnifPb_{i_l} \Ue \sigma_{i-1}\hoUnifPb_{i_r} \mapsto \sigma_i'$ and a link
  % progression $p = \progress(\linkStore_{i-1},\sigma_i') \mapsto (\linkStore_i,\sigma_i)$.
  We want to prove that \fstep succeeds iff \hstep
  succeeds, where \hstep is made by a unification step $u$ and a link
  progression $p$.
  Without loss of generality we consider a
  \Fo unification problem of the form $s_1 \Uo s_2$
  where $s_1 \in \maybeeta$ and is compiled to a variable $X$ with the accessory link
  $\linketaM{\Gamma}{X}{\lambda x.t_1}$. 
  The unification $u$ always succeeds, since $X$ is a fresh variable, we 
  only have to consider the link progression $p$.
  Two cases should be analyzed:
  \begin{itemize}
    \item $s_2 \notin \maybeeta$,
          in this case $s_2$ is compile to $t_2$ and the unification step $u$ assigns $t_2$ to $X$
          and triggers \progressetaleft.
          If $s_1$ and $s_2$ unify thanks to \ref{clause:eta1} then
          progression $p$ unifies $\lambda x.t_1$ with the $\eta$-expansion of $t_2$ namly
          $\lambda x.t_2\appsep x$, otherwise they unify thanks to \ref{clause:lam-lam}
          and the link progress $p$ unifies $\lambda x.t_1$ with $t_2$.\\ Alternatively, if $s_1$ and $s_2$ fail
          to unify, it cannot be because of the $\lambda$ constructor in the head of $t_1$,
          since \ref{clause:lam-lam} or \ref{clause:eta1} try to unify under it, eventually finding two
          different subterms $s_1'$ and $s_2'$. By \cref{th:fidelity-beta},
          \Ue fails on terms $t_1'$ and $t_2'$ even if \progressetaleft
          $\eta$-expands $t_2$.

@@

\begin{elpicode}
link-eta L R :- var L, maybe-eta R HolesInR, !,
  declare_constraint (link-eta L R) [L|HolesInR].
link-eta L R :- not (var L), !, eta-progress-lhs L R.
link-eta L R :- not (maybe-eta R), !, contract R S, L = S.
\end{elpicode}
\todo{check progress-rhs is back}
\noteD{first rule is wrong, since maybe-eta is 1-argument pred}
